---
title: "CMV primary episode analysis"
author: "Bryan Mayer"
date: "March 6, 2016"
output: pdf_document
toc: yes
---

```{r load knitr and set chunk options, echo = F}
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE,
              error = FALSE, tidy = FALSE)
```

```{r load packages, data, and functions}
library(plyr)
library(dplyr)
library(data.table)
library(knitr)
library(xtable)
library(ggplot2)
library(cowplot)

save_output_figs = F
save_output_tables = T
fontsize = 12
x_adj = 7
y_adj = 4

load("../data/CMVprimaryEpisode.RData")
source("functions/primary_episode_functions.R")

#to check and save appropriate tables (to avoid redoing bootstrap analysi)
results_files = list.files("results-data/")


episodeSummary = create_primary_episodes_fun(CMVPrimaryEpisodes)

process_stats_median = function(var){
  median_var = round(median(var), 2)
  range_var = round(range(var), 2)
  paste(median_var, " (", range_var[1], ", ", range_var[2], ")", sep = "")
}

process_est_ci= function(est, lower, upper){
  lower = round(lower, 2)
  upper = round(upper, 2)
  est = round(est, 2)
  paste(est, " (95% CI: ", lower, ", ", upper, ")", sep = "")
}

```

#Description

This is the main analysis of the primary CMV oral shedding episodes. Here, we extract key characteristics of the shedding episodes (e.g., peak load), determine the cutoffs for the three phases, analyze correlation between features, and analyze rebound behavior.

#Overview of the data

```{r tracings, fig.height=5, fig.width=7}
idorder = rev(levels(with(episodeSummary, reorder(PatientID2, -whenmax, mean))))
tracing_data = subset(CMVPrimaryEpisodes, count > 0)
tracing_data$PatientID2 = factor(tracing_data$PatientID2, levels = idorder, ordered = T)

patientSet = list(idorder[1:7], idorder[8:14])
tracings = llply(1:2, function(i){
  ggplot(data = subset(tracing_data, PatientID2 %in% patientSet[[i]]),
       aes(x = days2, y = count, colour = PatientID2)) +
    #geom_point(size = 1) +
    geom_line(size = 1) +
    scale_colour_discrete("Infant ID") +
    scale_x_continuous("Days from start of infection", breaks = 100*0:700, expand = c(0.01, 0.01), limits = c(0, 750)) +
    scale_y_continuous(expression(paste("Log"[10]," CMV DNA conc.")), breaks = 0:9, limits =  c(0, 9)) +
    theme(legend.position = "top",
      text = element_text(family = "Times", size = fontsize))
})

tracings_grid = plot_grid(tracings[[1]], tracings[[2]], nrow = 1,
                          labels = c("A", "B"), label_size = 18, vjust = 3.5)
tracings_grid
```

## General statistics, peak shedding, duration
```{r summary stats, results='asis'}

demo_stats = CMVPrimaryEpisodes %>% group_by(PatientID2, momhiv) %>% 
  summarize(age = min(days_dob),
            total_oral_swabs = n(),
            total_plasma_swabs = sum(!is.na(plasma_count)))

summary_statistics = data.table(
  n = length(unique(CMVPrimaryEpisodes$PatientID2)),
  'total negative swabs' = sum(CMVPrimaryEpisodes$count == 0),
  'median age (range) days' = process_stats_median(demo_stats$age),
  'percent mothers hiv +' = mean(demo_stats$momhiv == "pos") * 100,
  'total oral swabs' = sum(demo_stats$total_oral_swabs),
  'total plasma swabs' = sum(demo_stats$total_plasma_swabs),
  'median duration (range) days' = process_stats_median(episodeSummary$duration),
  'median peak shedding duration (range) weeks' = process_stats_median(episodeSummary$peak_shedding_days/7),
  'median first positive (range) log10 conc.' = process_stats_median(episodeSummary$firstpos),
  'median peak shedding (range) log10 conc.' = process_stats_median(episodeSummary$peak)
) %>% t()

colnames(summary_statistics) <- " "

print(xtable(summary_statistics, caption = "Summary statistics"), comment = F)

subset(CMVPrimaryEpisodes, count == 0) %>% group_by(PatientID2) %>% 
  summarize(n = n(), days = paste0(days2, collapse = ", ")) %>%
  xtable(caption = "Negative swabs") %>%
  print(include.rownames = F, comment = F)

```

```{r episode plot, fig.height=5.75, fig.width=4, fig.align='center'}
epPlot = ggplot(data = episodeSummary, aes(x = reorder(PatientID2, -whenmax, mean), y = whenmax)) +
  geom_linerange(aes(ymin = startTime, ymax = peakphase_start), 
                 size = 2) +
  geom_linerange(aes(ymin = peakphase_start, ymax = peak_shedding_days_end), 
                 size = 2, colour = "red") +
  geom_linerange(aes(ymin = peak_shedding_days_end, ymax = endTime), size = 2) +
  scale_x_discrete("Infant ID") +
  scale_y_continuous("Episode observation time (days)", breaks = 100 * 0:8) +
  geom_point(size = 4, colour = "red") +
  geom_point(aes(size = peak), shape = 21, colour = "red") +
  scale_size_continuous("Peak size", range = c(4, 13), breaks = c(7, 7.5, 8, 8.5, 9)) +
  #geom_rect(aes(xmin = as.numeric(PatientID) - 0.2, xmax = as.numeric(PatientID) + 0.2, 
  #              ymin = 0, ymax = 0.5)) +
  theme(
    text = element_text(family = "Times", size = fontsize)
    ,axis.text.y = element_text(size = fontsize)
    ,axis.title.y = element_text(size = fontsize)
    ,legend.position = c(1, 1)
    ,legend.justification = c(1, 1)) +
  coord_flip()

epPlot + ggtitle("Red region denotes shedding within 1 log of peak")
```

## Comparison of oral and plasma shedding

```{r plasma and oral plots, fig.height = 5.5, fig.width = 6, results = 'asis'}
plasma_data = subset(CMVPrimaryEpisodes, !is.na(plasma_count)) %>%
  mutate(positive_plasma = plasma_count > 0)

plasma_time_pl = ggplot(data = plasma_data, aes(x = days2, y = plasma_count)) +
  geom_point() +
  scale_x_continuous("Primary episode day") +
  scale_y_continuous("Log10 plasma CMV conc.")

plasma_oral_pl = ggplot(data = plasma_data, aes(x = count, y = plasma_count)) +
  geom_point() +
  scale_x_continuous("Log10 oral CMV conc.") +
  scale_y_continuous("Log10 plasma CMV conc.")

plot_grid(plasma_time_pl, plasma_oral_pl, nrow = 1)
```

```{r plasma and oral statistics}
pls_correlation_stat = with(plasma_data, cor.test(count, plasma_count)) %>%
  broom::tidy()


pls_correlation_stat_allpos = with(subset(plasma_data, positive_plasma), 
                                   cor.test(count, plasma_count)) %>%
  broom::tidy()

#spearman, use bootstrapping for CIs
pls_correlation_stat_spear = boot_fun(1000, select(plasma_data, count, plasma_count)) %>%
  rename(estimate = boot_median, conf.low = boot_lowerCI, conf.high = boot_upperCI)
pls_correlation_stat_allpos_spear = boot_fun(1000, select(subset(plasma_data, positive_plasma), count, plasma_count)) %>%
  rename(estimate = boot_median, conf.low = boot_lowerCI, conf.high = boot_upperCI)

pls_lmm = lmerTest::lmer(count ~ positive_plasma + (1|PatientID2), data = plasma_data)
pls_lmm_ci = confint(pls_lmm, parm = "positive_plasmaTRUE")

pls_lmm_results = data.frame(
  estimate = broom::tidy(pls_lmm) %>% filter(term == "positive_plasmaTRUE") %>% select(estimate),
  conf.low = as.numeric(pls_lmm_ci[1]),
  conf.high = as.numeric(pls_lmm_ci[2])
)
```

```{r plasma and oral results table, results = 'asis'}
plasma_table = data.table(
  'total oral swabs' = sum(demo_stats$total_oral_swabs),
  'total plasma swabs' = sum(demo_stats$total_plasma_swabs),
  'total positive plasma swabs' = sum(demo_stats$positive_plasma),
  'percent positive plasma swabs' = 100 * mean(plasma_data$positive_plasma),
  'median log10 pls viral load (range)' = 
    process_stats_median(subset(plasma_data, positive_plasma)$plasma_count),
  'Pearson correlation with oral' = with(pls_correlation_stat, process_est_ci(estimate, conf.low, conf.high)),
  'Spearman correlation with oral' = with(pls_correlation_stat_spear, 
                                   process_est_ci(estimate, conf.low, conf.high)),
  'Pearson correlation with oral (pos plasma only)' = 
     with(pls_correlation_stat_allpos, process_est_ci(estimate, conf.low, conf.high)),
  'Spearman correlation with oral (pos plasma only)' = with(pls_correlation_stat_allpos_spear, process_est_ci(estimate, conf.low, conf.high)),
  'lmm oral count ~ binary pls predictor' = 
    with(pls_lmm_results, process_est_ci(estimate, conf.low, conf.high))
) %>% t()
colnames(plasma_table) <- " "

print(xtable(plasma_table, caption = "Plasma load analysis"), 
      comment = F)
```


# Three phase classification
Each episode has three characteristic phases: 

1. expansion: oral episode start to 1 log before the peak

2. transition: 1 log before the peak until the peak

3. clearance: peak until end of observation 

```{r three phase setup for episodes}
phase_names = c("Expansion", "Transition", "Clearance")

durationData = episodeSummary %>% group_by(PatientID2) %>% 
  summarize(growth = peakphase_start,
            middle = peakphase_end - peakphase_start,
            decay = duration - peakphase_end) %>%
  melt(measure.vars = c("growth", "middle", "decay"))

swabVar = plyr::ldply(unique(CMVPrimaryEpisodes$PatientID2),
                function(id) swab2swab_function(subset(CMVPrimaryEpisodes, 
                                                       PatientID2 == as.character(id))))

triphasicFit = plyr::ldply(unique(CMVPrimaryEpisodes$PatientID2),
                     function(id) triphasicFit_function(subset(CMVPrimaryEpisodes, 
                                                               PatientID2 == as.character(id)))) %>%
  mutate(PatientID2 = as.character(PatientID2))

```
## Raw results
### Raw phase regression results
```{r raw regression results, results = 'asis'}
#Regression on each phase, note that this add confidence intervals
triphasicTable = plyr::ldply(unique(CMVPrimaryEpisodes$PatientID2), function(id)
    phase_regression_table_function(subset(CMVPrimaryEpisodes, PatientID2 == as.character(id))))

print(xtable(select(triphasicTable, PatientID2, expansion_n, 
                    expansion_slope_95pctCI, expansion_rsq, double_time_days), 
             caption = "Raw results from expansion phase regression"), 
      comment = F)
print(xtable(select(triphasicTable, PatientID2, trans_n, trans_slope_95pctCI, trans_rsq), 
             caption = "Raw results from transition phase regression"), 
      comment = F)
print(xtable(select(triphasicTable, PatientID2, clr_n, clr_slope_days_95pctCI,
                    half_life_days, clr_rsq), 
             caption = "Raw results from clearance phase regression"), 
      comment = F)

if(!any(grepl("phase_regression_results.csv", results_files))) readr::write_csv(triphasicTable,
                                                                      "results-data/phase_regression_results.csv")
```


## Summary of three phases

```{r three phase statistics, results = 'asis'}
#duration, slope, r2, swab to swab, exp(double/half)

#this is for transition
triphasicFit$temp = NA

feature_table = data.table(
  'Median duration (range) wk.' = 
    sapply(c("growth", "middle", "decay"), function(phase) 
      process_stats_median(subset(durationData, variable == phase)$value/7)),
  'Median swab change (range) wk.' = 
    sapply(c("growth", "middle", "decay"), function(ph) 
      process_stats_median(subset(swabVar, phase == ph & !is.na(change))$change)),
  'Median slope (range) wk.' = 
    sapply(c("growth", "middle", "decay"), function(phase) 
      with(triphasicFit, process_stats_median(get(phase) * 7))),
  'Median double/half-time (range) wk.' = 
    sapply(c("double", "temp", "half"), function(phase) 
      with(triphasicFit, process_stats_median(get(phase)/7))),
  'Median R-square (range)' = 
    sapply(c("rsq_1", "rsq_2", "rsq_3"), function(phase) 
      with(triphasicFit, process_stats_median(get(phase))))
) %>% t()
colnames(feature_table) = c("Expansion", "Transition", "Clearance")

print(xtable(feature_table, caption = "Summary of phase features"), 
      comment = F)

cat("Infant", triphasicFit$PatientID2[which.min(triphasicFit$rsq_1)], "had the lowest expansion R2 of", min(triphasicFit$rsq_1), ".\n")
low_clear = which(triphasicFit$rsq_3<0.5)
for(i in 1:length(low_clear)){
  cat("Infant", triphasicFit$PatientID2[low_clear[i]], "had the a low clearance R2 of", triphasicFit$rsq_3[low_clear[i]], ".\n")
}
```

## Compare three phase to two phase
The three phase classification comes from an apparent transition phase between the start of the episode and the peak viral load. In the biphasic classification, expansion phase ends at peak (instead of 1 log before peak). The R-square values are generally worse when we make that assumption.

```{r, results = 'asis'}

biphasicFit = plyr::ldply(unique(CMVPrimaryEpisodes$PatientID2),
                     function(id) biphasicFit_function(subset(CMVPrimaryEpisodes, 
                                                               PatientID2 == as.character(id))))

compare_expansion = full_join(select(triphasicFit, PatientID2, rsq_1),
                              select(biphasicFit, PatientID2, rsq_peak),
                              by = "PatientID2") %>%
  mutate(better = rsq_peak < rsq_1, rsq_1 = round(rsq_1, 2))


compare_expansion = with(compare_expansion, data.table(
  'Infant ID' = PatientID2,
  'Growth to peak' = rsq_peak,
  'Growth to 1 log before peak' = rsq_1,
  better = better,
  'Triphasic > biphasic' = better
))

print(xtable(select(compare_expansion, -better), caption = "R-square comparing growth in two phase vs three phase classification"), include.rownames = F, comment = F)

cat(sum(compare_expansion$better), "episodes perform better with transition phase")

```


## Plot of features
```{r plots of phase characteristics, fig.width = 12, fig.height = 10, fig.pos = 't', results = 'asis'}

##### first pos vs peak ########
firstposPeak = ggplot(data = episodeSummary %>%
                        reshape2::melt(measure.var = c("firstpos", "peak"),
                                            value.name = "count", variable.name = "time"),
                      aes(x = time, y = count)) +
  geom_boxplot(aes(fill = time)) +
  scale_fill_manual(guide = F,
                    values = gray.colors(3, start = 1, end = 0.5, gamma = 2.2, alpha = 0.5)) +
  scale_x_discrete("", labels = c("First Positive", "Peak")) +
  scale_y_continuous(expression(paste("log"[10]," CMV DNA conc.")), breaks = 1:10) +
  theme(
    text = element_text(family = "Times", size = fontsize),
    axis.title.y = element_text(size = fontsize - y_adj)
  )

##### Duration ####
duration_pl = ggplot(data = subset(durationData, !(variable == "decay" & value == 0)), 
                     aes(x = variable, y = value/7)) +
  geom_boxplot(aes(fill = variable)) +
  scale_fill_manual(guide = F, breaks =  c("growth", "middle", "decay"),
                    values = c("White", "#CCCCCC80", "#80808080")) +
  scale_x_discrete("Episode phase", limits = c("growth", "middle", "decay"),
                   labels = c(phase_names[1:2], "Clearance\n(Censored)")) +
  scale_y_continuous("Duration (wks)", breaks = 10*0:10)+
  theme(
    axis.text.x = element_text(size = fontsize - x_adj),
    axis.title.x = element_text(size = fontsize - 5),
    axis.title.y = element_text(size = fontsize - y_adj),
    text = element_text(family = "Times", size = fontsize))

####### swab to swab variability ########
swab_pl = ggplot(data = swabVar, aes(x = phase, y = change)) +
  geom_boxplot(aes(fill = phase)) +
  scale_fill_manual(guide = F, breaks =  c("growth", "middle", "decay"),
                    values = c("White", "#CCCCCC80", "#80808080")) +
  scale_x_discrete("Episode phase", limits = c("growth", "middle", "decay"), labels = phase_names) +
  scale_y_continuous(expression(paste("Swab to swab change(log"[10]," CMV DNA conc.)")),
                     limits = c(-3, 3), breaks = -6:6) +
  geom_hline(aes(yintercept = 0), alpha = 0.5, linetype = "dashed") +
  geom_text(aes(x = phase, label = count, y = -Inf), vjust = -0.3, 
            data = swabVar %>% group_by(phase) %>%
              dplyr::summarize(count = n()))+
  theme(
    axis.text.x = element_text(size = fontsize - x_adj),
    axis.title.x = element_text(size = fontsize - 5),
    axis.title.y = element_text(size = fontsize - y_adj),
    text = element_text(family = "Times", size = fontsize))

##### Slope #####

slope_pl = ggplot(data = triphasicFit %>%
                    reshape2::melt(measure.vars = c("growth", "middle", "decay"),
                                   value.name = "slope", variable.name = "phase"), 
                  aes(x = phase, y = slope*7)) +
  geom_boxplot(aes(fill = phase)) +
  scale_fill_manual(guide = F, breaks =  c("growth", "middle", "decay"),
                    values = c("White", "#CCCCCC80", "#80808080")) +
  scale_x_discrete("Episode phase", limits = c("growth", "middle", "decay"), labels = phase_names) +
  scale_y_continuous(expression(paste("Estimated weekly change(log"[10]," CMV DNA conc./wk)")),
                     limits = c(-0.5, 2), breaks = -1:4/2) +
  geom_hline(aes(yintercept = 0), alpha = 0.5, linetype = "dashed") +
  theme(
    axis.text.x = element_text(size = fontsize - x_adj - 0.5),
    axis.title.y = element_text(size = fontsize - y_adj),
    axis.title.x = element_text(size = fontsize - 5),
    text = element_text(family = "Times", size = fontsize))

##### Rsq ######
rsq_pl =  ggplot(data = triphasicFit %>%
                   reshape2::melt(measure.vars = c("rsq_1", "rsq_2", "rsq_3"),
                                  value.name = "rsq", variable.name = "rsqphase"),  
                 aes(x = rsqphase, y = rsq)) +
  geom_boxplot(aes(fill = rsqphase)) +
  scale_fill_manual(guide = F, breaks =  c("growth", "middle", "decay"),
                    values = gray.colors(3, start = 1, end = 0.5, gamma = 2.2, alpha = 0.5)) +
  scale_x_discrete("Episode phase", limits = c("rsq_1", "rsq_2", "rsq_3"), labels = phase_names) +
  scale_y_continuous(expression(paste(R^2)), limits = c(0, 1), breaks = -1:4/2) +
  theme(
    axis.text.x = element_text(size = fontsize - x_adj),
    axis.title.x = element_text(size = fontsize - 5),
    axis.title.y = element_text(size = fontsize - y_adj),
    text = element_text(family = "Times", size = fontsize))

##### Doubling and half-life #####
dbl_pl = ggplot(data = triphasicFit %>%
                  reshape2::melt(measure.vars = c("double", "half"),
                                 value.name = "dbl", variable.name = "phase"),
                aes(x = phase, y = dbl)) +
  geom_boxplot(aes(fill = phase)) +
  scale_fill_manual(guide = F, breaks =  c("double", "half"),
                    values = gray.colors(3, start = 1, end = 0.5, gamma = 2.2, alpha = 0.5)[c(1,3)]) +
  scale_x_discrete("", limits = c("double", "half"), labels = c("Doubling time", "Half-life")) +
  scale_y_continuous("Days", breaks = 10 * 0:10) +
  theme(
    axis.title.y = element_text(size = fontsize - y_adj),
    text = element_text(family = "Times", size = fontsize))

phase_feature_plot = plot_grid(firstposPeak, duration_pl, swab_pl, slope_pl, dbl_pl, rsq_pl,
                    labels = c("A", "B", "C", "D", "E", "F"), nrow = 2,
                    label_size = 18, hjust = 0)

phase_feature_plot

```


## Cubic model of clearance and rebound analysis

```{r cubic analysis, fig.align = 'center', fig.height=4}

cubic_fit = plyr::ldply(unique(CMVPrimaryEpisodes$PatientID2), function(pid) cubic_decay_fit(pid))

ggplot(data = subset(CMVPrimaryEpisodes, PatientID2 %in% c("H", "L") & count > 0), aes(x = days2, y = count2)) +
  geom_line() + geom_point() + facet_wrap(~PatientID2, nrow = 2) +
  ggtitle("Borderline deceleration cases")

cubic_fit$critPoint2[is.na(cubic_fit$critPoint2)] = Inf

cubic_fit$rebound_cat = with(cubic_fit, ifelse(critPoint2 < last_day, 2,
                                               ifelse(last_day - inf_point < 100 | !is.na(minPoint2), 1, 0)))
cubic_fit$frompeak = with(cubic_fit,
                             ifelse(rebound_cat == 2, critPoint2 - peak_day, last_day - peak_day))

minEpCrit = with(subset(cubic_fit, rebound_cat == 2), critPoint2[which(critPoint2 == min(critPoint2))])
maxEpCrit = with(subset(cubic_fit, rebound_cat == 2), critPoint2[which(critPoint2 == max(critPoint2))])
minPeakCrit = with(subset(cubic_fit, rebound_cat == 2), frompeak[which(frompeak == min(frompeak))])
maxPeakCrit = with(subset(cubic_fit, rebound_cat == 2), frompeak[which(frompeak == max(frompeak))])

cubic_fit$max_day = with(cubic_fit, ifelse(rebound_cat == 2, critPoint2, last_day))

rebound_pl = ggplot(data = cubic_fit, aes(x = max_day, y = frompeak)) +
  geom_vline(aes(xintercept = minEpCrit), linetype = 1) +
  geom_vline(aes(xintercept = maxEpCrit), linetype = 1, colour = "red") +
  geom_hline(aes(yintercept = minPeakCrit), linetype = 2) +
  geom_hline(aes(yintercept = maxPeakCrit), linetype = 2, colour = "red") +
  geom_point(aes(shape = factor(rebound_cat)), size = 10, position = position_dodge(width = 0.25), fill = "#80808080") +
  scale_shape_manual("", values = c(21, 24, 22),
                     breaks = 0:2, labels = c("No rebound", "Deceleration",
                                              "Rebound")) +
  scale_y_continuous("Days from start of clearance (peak)", breaks = 100*0:7, limits = c(0, 675)) +
  scale_x_continuous("Days from start of infection", breaks = 100*0:8, limits = c(200, 775)) +
  geom_text(aes(label = PatientID2), size = 7) +
  theme(
    text = element_text(family = "Times"),
    legend.position = "top",
    legend.key = element_blank())

rebound_pl

```

## Correlation analysis

### Age and features

```{r correlation analysis, results = 'asis', message=F}

correlation_data = select(triphasicFit, PatientID2, age, peak, first_peak_day, high_shed_length, decay_length)

#names(compare_data) = c("Age (days)", "Peak", "Growth duration\n(days)", "Transition\nduration (days)", "Clearance\nduration (days)")
variable_set = c("age", "peak", "first_peak_day", "high_shed_length", "decay_length")
names_set =  c("Age (days)", "Peak", "Expansions days", "Transition days", "Clearance days")
rename_set = data.frame(var_name = variable_set, label_name = names_set, stringsAsFactors = F)

all_pairs = combn(variable_set, 2)

if(!any(grepl("bootstrap_results_spearman.csv", results_files))){
  library(doParallel)
  registerDoParallel(cores = 8)
  
  boot_samples = ldply(1:dim(all_pairs)[2], function(pair){
    pair_data = select(correlation_data, get(all_pairs[1, pair]), get(all_pairs[2, pair]))
    boot_fun(10000, pair_data)
  }, .parallel = T)
  readr::write_csv(boot_samples, "results-data/bootstrap_results_spearman.csv")
  
} else boot_samples = readr::read_csv("results-data/bootstrap_results_spearman.csv")


if(!any(grepl("bootstrap_results_noageoutliers_spearman.csv", results_files))){
  library(doParallel)
  registerDoParallel(cores = 8)  
  
  boot_samples_ageoutliers = ldply(1:dim(all_pairs)[2], function(pair){
    pair_data = select(subset(correlation_data, age < 150), get(all_pairs[1, pair]), get(all_pairs[2, pair]))
    boot_fun(10000, pair_data)
  }, .parallel = T)
  readr::write_csv(boot_samples_ageoutliers, "results-data/bootstrap_results_noageoutliers_spearman.csv")
  
} else boot_samples_ageoutliers = readr::read_csv("results-data/bootstrap_results_noageoutliers_spearman.csv")


process_correlation_results = function(raw_boot_in){
  raw_boot_in$pair1 = sapply(raw_boot_in$pair1, function(pair_name) rename_set$label_name[rename_set$var_name == pair_name],
                             USE.NAMES = F)
  raw_boot_in$pair2 = sapply(raw_boot_in$pair2, function(pair_name) rename_set$label_name[rename_set$var_name == pair_name],
                             USE.NAMES = F)
  raw_boot_in = select(raw_boot_in, pair1, pair2, base_cor, base_pvalue, boot_median, boot_lowerCI, boot_upperCI)
  names(raw_boot_in) <- c("Pair1", "Pair2", "raw_cor", "p_value", "boot_cor", "boot_lower95CI", "boot_upper95CI")
  raw_boot_in
}

print(xtable(process_correlation_results(boot_samples_ageoutliers), digits = 3, 
             caption = "Spearman Correlations - excluded 2 older infants. Bootstrap takes median."), comment = F, include.rownames = F)

print(xtable(process_correlation_results(boot_samples), digits = 3, 
             caption = "Spearman Correlations - all data. Bootstrap takes median."),
      comment = F, include.rownames = F)


```

### Mother HIV

```{r momhiv correlations, results = 'asis'}
momhiv_ttests = plyr::ldply(variable_set, function(var){
  reg_formula = as.formula(paste(var, "~momhiv"))
  output = broom::tidy(lm(reg_formula, data = triphasicFit)) %>%
    mutate(outcome = rename_set$label_name[rename_set$var_name == var])
  select(subset(output, term != "(Intercept)"), outcome, estimate, p.value)
})


print(xtable(momhiv_ttests, digits = 3, 
             caption = "Linear model (unadjusted) with momhiv (reference = neg) as predictor."), 
      comment = F, include.rownames = F)

if(!any(grepl("momhiv_results.csv", results_files))) readr::write_csv(momhiv_ttests,
                                                                      "results-data/momhiv_results.csv")

```


# Saving figures and table 

See code, won't save new tables or figures unless the save_output_figs and save_output_tables variables are changed to TRUE.

## Figures

```{r saving figures}
if(save_output_figs){
  
  ##### Figure 1 ########
  # tracings_grid and ep_plot comes from Overview of the Data section
  example_ep = ggplot(data = subset(CMVPrimaryEpisodes, PatientID2 == "C" & count > 0),
                      aes(x = days2, y = count)) +
    annotate("rect", xmin = -Inf, xmax = 70, ymin = -Inf, ymax = Inf, 
             fill = "#FFFFFF", colour = NA, alpha = 0.4) +
    annotate("rect", xmin = 70, xmax = 147, ymin = -Inf, ymax = Inf, 
             fill = "#CCCCCC", colour = NA, alpha = 0.4) +
    annotate("rect", xmin = 147, xmax = Inf, ymin = -Inf, ymax = Inf, 
             fill = "#808080", colour = NA, alpha = 0.4) +
    geom_point(size = 4) +
    geom_line(size = 1.5) +
    scale_x_continuous("Days from start of infection",breaks = 100*0:600, expand = c(0.01, 0.01)) +
    scale_y_continuous(expression(paste("Log"[10]," CMV DNA conc.")), breaks = 0:9) +
    theme(
      text = element_text(family = "Times", size = fontsize))

  ep_grids = plot_grid(epPlot, example_ep, nrow = 1,
          labels = c("C", "D"), label_size = 18)

  figure1 = plot_grid(tracings_grid, ep_grids, nrow = 2)
  ggsave("../figures/Figure1.pdf",
         figure1,
         width = 11.5, height = 12)
  
  ##### Figure 2 #########
  # The 6 plots comes from Three Phase Classification -> Plot of Features subsection
  
  figure2 = plot_grid(firstposPeak, duration_pl, swab_pl, slope_pl, dbl_pl, rsq_pl,
                    labels = c("A", "B", "C", "D", "E", "F"), label_size = 18, hjust = 0)
  ggsave("../figures/Figure2.pdf",
         figure2,
         width = 12, height = 10)
  
  ###### Supplementary figure for rebound
  # Data comes from cubic rebound subsection
  
  ggsave("../figures/supplementary-figures/suppfig_cubic_rebound.pdf",
         rebound_pl, width = 11.5, height = 9)

}


```

```{r saving tables}
if(save_output_tables){
  library(readr)
  library(WriteXLS)
  
  #Demographic and phase features
  demo_table = triphasicFit[, c("PatientID2", "age", "momhiv", "peak", 
                                "first_peak_day", "high_shed_length", "decay_length", "total_days" )]

  #these names come from old iterations, easier to change them here to potentially break code and 
  # change in the functions scripts
  
  names(demo_table) = c("PatientID2", "age", "momhiv", "peak", 
                                "expanstion_length", "transition_length", "clearance_length", "total_days" )
  write_csv(demo_table, "results-data/demo_features_data.csv")
  
  
  # make them look a little nicer and save together in an excel spreadsheet
  for(i in 1:length(colnames(demo_table))){
    if(colnames(demo_table)[i] == "peak") demo_table[, i] = round(demo_table[, i], 2)
    else demo_table[, i] = as.character(demo_table[, i])
  }
  names(demo_table) <- c("Infant ID", "Age (days)", "Mother HIV", "Peak viral load (log10 DNA copies/swab)",
                         "Expansion duration (days)", "Transition duration (days)", "Clearance duration (days)",
                         "Episode obs. time (days)")
  
  regression_table = triphasicTable
  names(regression_table) = c("Infant ID", "Expansion N", "Expansion slope (/day) (95% CI)", "Expansion R2",
                              "Doubling time (days)", "Transition N", "Transition slope (/day) (95% CI)",
                              "Transition R2", "Clearance N", "Clearance slope (/day) (95% CI)", 
                              "Clearance R2", "Half-life (days)")
  
  phase_results_table = full_join(demo_table, regression_table, by = "Infant ID")
  
  WriteXLS(phase_results_table, "../tables/supplementary-tables/primary_episode_analysis_results.xls")
  
  
  #Supplementary Cubic table
  cubic_table =  plyr::ldply(unique(CMVPrimaryEpisodes$PatientID), function(pid) cubic_decay_fit_table(pid))
  write_csv(cubic_table, "../tables/supplementary-tables/supptable_cubic_regression.csv")

}

```