---
title: "parameter_boundary"
output: pdf_document
---
This function is sourced and called throughout the optimization workflow for boundaries.  It contains all the parameters I used.
MAKE SURE DOCUMENTATION MATCHES THE ACTUAL SET VALUES
(make sure listed parameter ranges are reflected in the function)

# Functions for parameters

Biological information, parameter interpretation given at the end. Also a calculation to draw ranges of $\beta$ are shown but the model constraints are set to be much wider.

## Boundary Function used in the optimizations -- all in log$_{10}$
```{r}
#### EDIT THIS IN THE DOCUMENTATION FILE (.Rmd) ONLY ####


boundary_set = function(parms, boundary = NULL, t_set = NA, par_set = NA, initV_upper = 2){
  if(length(parms) == 0 | is.null(boundary)) return(NULL)
  
  #remember these are all log10ed
  
  if(boundary %in% c("l", "lower", "L", "Lower")){
    set_values = c(
      beta = -14,
      delta = -4,
      theta = -3,
      KI = 0,
      KT = 2,
      death = -5,
      start_day = log10(0.1)
    )
    return_list = sapply(1:length(parms), function(i) which(names(set_values) == names(parms)[i]))
    
    return(unname(set_values[return_list]))
  }
  
  if(boundary %in% c("u", "upper", "U", "Upper")){
    set_values = c(
      beta = -6.5,
      delta = 3,
      theta = 2,
      KI = 8,
      KT = 10,
      death = 0,
      start_day = log10(25)
    )
    return_list = sapply(1:length(parms), function(i) which(names(set_values) == names(parms)[i]))
    
    return(unname(set_values[return_list]))
  }
  
  print("Incorrect boundary label given (use 'lower' or 'upper')")
  return(NULL)
}
```

## Latin hypercube initial value draw functions

R0 checking (constraining the sampling range) is better handled directly in the specific optimization scripts.

```{r}

#this returns a list of a set of initial value draws for given set of parameters
#uses LHS package
#R0check is only for when beta and delta are fit together
draw_initial = function(fitparms, lower, upper,
                        total_draws = 50, save.out = T, R0check = F, file_name = NULL){
  totalFit = length(fitparms)
  
  if(totalFit != length(lower) | totalFit != length(upper)) {
    print(paste("mismatch, 1=lower, 2=upper:", which(!c(length(lower), length(upper)) %in% totalFit)))
    return(NULL)
  }
  
  
  lhs = randomLHS(total_draws, totalFit)
  
  starting_sets = llply(1:total_draws, function(i){
    tempdraws = (upper - lower) * lhs[i, ] + lower
    names(tempdraws) = names(fitparms)
    
    if(R0check){
      tempR0 = with(as.list(10^tempdraws), beta * 4e8 * 60 /(delta * 2))
      if(tempR0 < 0.5) return(NULL)  
    }

    tempdraws
  })
  
  starting_sets <- compact(starting_sets) #compact removes NULLs (remove nothing if R0check = F)
  total_kept = length(starting_sets) #should be same as total_draw when R0check = F
  
  if(save.out){
    starting_sets_save = ldply(1:total_kept, function(i) starting_sets[[i]])
    if(is.null(file_name)) file_name = Sys.Date()
    write.csv(starting_sets_save,
            paste(file_name,".csv", sep = ""),
            row.names = F)
  }
  
  return(starting_sets)

}
```


## addnames_fit

```{r}

#this tacks on parameters names that werent used in the model for rbinding in make_plot_data and main ldply
addnames_fit = function(output, parmnames){
  addnames = which(!parmnames %in% names(output))
  
  if(length(addnames) == 0) return(output)
  
  temp_output = cbind(output, t(rep(NA, length(addnames))))
  names(temp_output) = c(names(output), parmnames[addnames])
  
  return(temp_output)
  
}
```


# Background information

## Equations

### 1. SIV model with no immunity

$$\frac{\mathrm{d}S}{\mathrm{d}t} = \lambda - \mu S - \beta S V $$
$$\frac{\mathrm{d}I0}{\mathrm{d}t} = \beta S V - \alpha I0 - \mu I0 $$
$$\frac{\mathrm{d}I}{\mathrm{d}t} = \alpha I0 - \delta I - kIT $$
$$\frac{\mathrm{d}V}{\mathrm{d}t} = pI - cV$$
$$ \frac{\mathrm{d}T}{\mathrm{d}t} = \theta \frac{I}{K_I + I} - death*T $$

### 2. Model with cytolytic immunity

$$\frac{\mathrm{d}S}{\mathrm{d}t} = \lambda - \mu S - \beta S V $$
$$\frac{\mathrm{d}I0}{\mathrm{d}t} = \beta S V - \alpha I0 - \mu I0 $$
$$\frac{\mathrm{d}I}{\mathrm{d}t} = \alpha I0 - \delta I - kIT $$
$$\frac{\mathrm{d}V}{\mathrm{d}t} = pI - cV$$
$$ \frac{\mathrm{d}T}{\mathrm{d}t} = \theta \frac{I}{K_I + I} - death*T $$

### 3. Model with viral-mediated response

$$\frac{\mathrm{d}S}{\mathrm{d}t} = \lambda - \mu S - \beta S V $$
$$\frac{\mathrm{d}I0}{\mathrm{d}t} = \beta S V - \alpha I0 - \mu I0 $$
$$\frac{\mathrm{d}I}{\mathrm{d}t} = \alpha I0 - \delta I $$
$$\frac{\mathrm{d}V}{\mathrm{d}t} = pI - cV - kTV$$
$$\frac{\mathrm{d}T}{\mathrm{d}t} = \theta \frac{V}{K_T + V} - death*T $$


## Biologically fixed parameter interpretation and ranges
$\lambda = K\mu$ -- Fixed growth rate of S fixed

$K = 10^7 * 40$ Fixed total S size (from literature, Dawes 2003)

$\mu = 1/4.5$ -- Death rate of S or I0 (also from Dawes)

$\beta\in 10^{-14}, 10^{-6.5}$ -- Infectivity, see below for constraint calculations.

$\alpha = 1$ -- Latency period before virus replicates

$\delta = 0.77$ -- Death rate of I, per day (Emery paper)

$\delta\in (10^{-4}, 10^{3})$ -- These boundaries were used when profiling over $\mu$. 

$p = 1600$ -- PNAS temperature paper.

$c = 2$ -- Viral clearance rate (generally fixed), 2 from EBV model paper

$k = 0.01$ -- Immune clearance of virus (or maybe I), fixed from several papers at 1% per effectory per day.

$\theta\in (0.001, 100)$ -- Immune activation rate

$K_I\in (100, 10^{8})$ -- Infected cell population where immune system is 50% activated

$death\in (0.00001, 1)$ -- Death rate of immune effector population

$K_T\in (100, 10^{10})$ -- Viral level where immune system is 50% activated

start_day or $t_0\in (0.1, 25)$ -- Days before detection of first positive. The day when $I_0 = 1$ in the model simulation.

## Constraining $\beta$ for sampling

$$ R_0 = \frac{\beta p\lambda}{c\delta\mu (1+\frac{\mu}{\alpha})} =  \frac{\beta p K}{c\delta (1+\frac{\mu}{\alpha})}$$

$$ 1 < R_0 < X $$

so the general solution for $\beta$ here is

$$\frac{c\delta(1+\frac{\mu}{\alpha})}{p K} < \beta < X*\frac{c\delta(1+\frac{\mu}{\alpha})}{p K}$$

$\beta$ was sampled from this range for target cell model. When $\mu$ was varied and $\delta$ optimized with $\beta$, LHS sampled values were tested to make sure R0 was between 1 and 50.