---
title: "Analysis of cytolyic immunity models"
output: pdf_document
toc: yes
fig_caption: yes
---
# Overview

This document goes over the analysis of the model results from the main model described in the manuscript: CMV infection with cytolytic immune response (see model_code_documentation/CMV_Models.pdf for equations and code). The immune parameters for this model are fit across the entire episode for each infant. The beta and start_time parameters were fit to the expansion phase only in a prior analysis to estimate R0 (see first results section of target_cell_model_analysis.pdf for that).

This document and code generated the results presented for publication: Figures 3 and 4 come from here. This is how the analysis was broken down:

1. Assessement of optimized parameter values and model fit
2. Model results interpretation through simulation using best fits and estimation of infected cell lifespan with immune response
3. Estimation of effective reproduction number, R, to evaluate critical time points during infection
4. Check on epithelial cell loss


```{r, warning = FALSE, message = FALSE, echo = F}
library(dplyr)
library(readr)
library(deSolve)
library(ggplot2)
library(grid)
library(gridExtra)
library(scales)
library(RColorBrewer)
library(GGally)
library(cowplot)
library(xtable)
library(knitr)
library(gtable)
library(doParallel)
registerDoParallel(cores = 4)
theme_set(theme_bw())

save_output_figs = F
save_output_tables = F

fontsize_set = 13
x_lims = c(0, 425)


opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE,
              error = FALSE)

theme_set(theme_bw())

source("model_functions/CMV_Models.R")
source("model_functions/processing_functions.R")
source("parameters/parameter_boundary.R")
load("../data/CMVprimaryEpisode.RData")
```


# Model optimization and results

Model parameters were estimated for each infant episode after initializing. Subject by subject parameters and model fits appear in the results_figures/

```{r load and process data, warning = FALSE, message = FALSE, echo = F}
immune_model_raw = read_csv("model_results/cytolytic_immune_model_fits.csv")
immune_model_raw$mse2 = immune_model_raw$mse + 1e-9*runif(length(immune_model_raw$mse))
immune_model_fits = immune_model_raw %>% group_by(PatientID2, model) %>% dplyr::mutate(best = (mse2 == min(mse2))) %>%
  filter(best == 1) %>% select(-mse2, -best) %>% ungroup()
immune_model_fits$betaK = with(immune_model_fits, beta * K)

parameter_list = c("theta", "KI", "death", "mse")

```

## Fitted model parameters
```{r,  message = F, results='asis'}

save(immune_model_fits, file = "model_results/ctl_immune_model_fits.RData")

immune_fits_table = immune_model_fits %>%
  mutate(log10KI = log10(KI), gammaX100 = death * 100) %>%
  select(PatientID2, theta, log10KI, gammaX100, mse)

print(xtable(immune_fits_table,
                    caption = "Supplementary Table - Fits from cytolytic immune model"),
      comment = F)


```

```{r, message = F, fig.cap='Parameter boxplots'}
parameter_plot(immune_model_fits, parameter_list) +
  theme(legend.position = "none") +
  ggtitle(immune_model_fits$model[1])
```


```{r, message = F, fig.cap='Distribution and correlation of fitted immune parameters (all transformed log10)'}
print(ggpairs(log10(subset(immune_model_fits, model == "CMVModel_latent_immunity_CTL")[, parameter_list])),
      bottomHeightProportion = 0.3, leftWidthProportion = 0.3, spacingProportion = 0.1)
```

## Simulated data fit compared to episode data (Figure 3)

(also saves figure results_figures/CTL_model_fits.pdf that displays parameter values, R0 along with time series plots. This chunk has eval = F in the code.)

```{r simulate results, message = FALSE, echo = F, eval = T}

sim_data = make_subject_plots(immune_model_fits, CMVPrimaryEpisodes, return_data = T, parallel = T, first_time = T)
sim_data$model = as.character(sim_data$model)
```


```{r save parm results, message = FALSE, echo = F, eval = F}
### only need to run this chunk if you want to save new model fit plots

parmplots = data_parameter_plot(immune_model_fits, sim_data, parameter_list)
out_parmplots = marrangeGrob(grobs = parmplots, nrow = 1, ncol = 1)

ggsave("results_figures/CTL_model_fits.pdf", out_parmplots)


```

```{r plot model fits, fig.align ='center', fig.cap='Best immune model fits', fig.height = 9, fig.width = 5}


imm_plots = plyr::llply(unique(sim_data$PatientID2), function(pid){
  temp_data = subset(sim_data, PatientID2 == pid)
  ggplot() +
    geom_point(data = subset(temp_data, model == "data"), aes(x = days_model, y = count), colour = "black") +
    geom_line(data = subset(temp_data, model != "data"),
              aes(x = days_model, y = count, colour = model, linetype = model), size = 1.5, alpha = 0.7) +
    scale_x_continuous("", breaks = 100*0:8) +
    scale_y_continuous("", breaks = 2*0:5, limits = c(0, 10.5)) +
    scale_colour_discrete("") +
    scale_linetype_discrete("") +
    theme(legend.position = "none",
          plot.margin = unit(c(0, 0.2, 0, -0.1),"cm"),
          text = element_text(family = "Times"),
          title = element_text(size = 9.5),
          legend.position = "none") +
    ggtitle(temp_data$PatientID2[1])

}, .parallel = F)



bot_lab = "Time (days)"
left_lab = expression(paste("Log"[10]," CMV DNA conc."))

all_plots = arrangeGrob(grobs = imm_plots, nrow = 7, ncol = 2,
                        textGrob(bot_lab, vjust = -0.2, gp=gpar(fontfamily = "Times", fontsize = 18)),
                        left = textGrob(left_lab, vjust = 1.2, rot = 90, gp=gpar(fontfamily = "Times", fontsize = 18)))

plot_grid(all_plots)


fit_plots = ggplot() +
  geom_point(data = subset(sim_data, model == "data"), aes(x = days_model, y = count), colour = "black") +
  geom_line(data = subset(sim_data, model != "data"),
          aes(x = days_model, y = count, colour = model, linetype = model), size = 1.5, alpha = 0.7) +
  scale_x_continuous("Time (days)", breaks = 100*0:8) +
  scale_y_continuous("", breaks = 2*0:5, limits = c(0, 10.5)) +
  scale_colour_discrete("") +
  scale_linetype_discrete("") +
  facet_wrap(~PatientID2, ncol = 2, scales = "free_x") +
  theme(legend.position = "none",
          plot.margin = unit(c(0, 0.2, 0, -0.1),"cm"),
          text = element_text(family = "Times"),
          title = element_text(size = 9.5),
          legend.position = "none") 

  ggsave("../figures/Figure3TEST.pdf",
     fit_plots, width = 210, height = 297, units = "mm")
```


# Infected cell lifespan with immune pressure

## Time series plots of immune pressure
```{r calculate immune pressure and effR}
immune_sim_only = merge(subset(sim_data, model != "data"), immune_model_fits, by = c("model", "PatientID2"), all.x = T) %>%
  group_by(PatientID2) %>% mutate(start_day = abs(min(days_model))) %>%
  ungroup() %>%
  mutate(
    days_model_adj = days_model + start_day, #this is to rescale the simulation to be day from start of infection (instead of first positive)
    clr = 0.01 * Tcell + 0.77,
    action = (0.77 + 0.01 * Tcell)/24,
    action_plot = 1/action,
    action_pct = action_plot/max(action_plot) * 100,
    effR0 = S/K * R0 * delta / (delta + k * Tcell),
    effR0_noS = R0 * delta / (delta + k * Tcell)
  )


immune_summary_data = immune_sim_only %>% group_by(PatientID2, R0) %>%
  summarize(
    minimum_lifespan_hrs = action_plot[which(Tcell == max(Tcell))],
    maximum_reduction_pct = 100 - action_pct[which(Tcell == max(Tcell))],
    peak_dayV = days_model_adj[which(V == max(V, na.rm = T))],
    peak_dayI = days_model_adj[which(I == max(I))],
    peak_dayI0 = days_model_adj[which(I0 == max(I0))],
    peak_immune_day = days_model_adj[which(Tcell == max(Tcell))],
    cross_day = days_model_adj[min(which(round(effR0, 3) == 1))], # should be same as peak_dayV
    minR = min(effR0) # remember that minimum R comes from both susceptible cell depletion and immune response
    )
 
peak_time_table = immune_summary_data

```

```{r time series immune pressure, fig.align = 'center', fig.cap='Immune pressure over time on infected cell lifespan', fig.height = 4, fig.width = 4}
dyn_plot = ggplot(data = immune_sim_only, aes(x = days_model, y = 100 - action_pct)) +
  geom_line(aes(colour = PatientID2)) +
  scale_y_continuous("Infected cell lifespan reduction (%)", limits = c(0, 75), breaks = 0:4 * 25) +
  scale_x_continuous("Time (days)", limits = x_lims, breaks = 100 * 0:6) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        axis.title.y = element_text(size = 11),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times")
  )

dyn_plot
```


## Magnitude of peak response (shortest infected cell lifespan)

```{r, results = 'asis'}

print(xtable(
  select(peak_time_table, PatientID2, minimum_lifespan_hrs, maximum_reduction_pct), 
  digits = 2, caption = "Maximum immune pressure effect on infected cell lifespan (compared to 31 hours without immunity)"),
      comment = F, include.rownames = F)

peak_time_table %>% ungroup() %>% reshape2::melt(measure.vars = c("minimum_lifespan_hrs", "maximum_reduction_pct")) %>%
  group_by(variable) %>%
  summarize(
    median = median(value),
    IQR = paste(round(quantile(value, c(0.25, 0.75)), 3), collapse = ", "),
    range = paste(round(range(value), 3), collapse = ", ")
  ) %>%
  xtable(caption = "Summary of maximum immune pressure effect on infected cell lifespan  (compared to 31 hours without immunity)") %>%
  print(comment = F, include.rownames = F)

```

```{r magnitude of peak response, fig.align='center', fig.height = 3, fig.width = 3}
peak_box = ggplot(data = immune_summary_data, aes(x = "", y = minimum_lifespan_hrs)) +
  geom_boxplot(width = 1, outlier.colour = alpha("white", 0)) +
  geom_point(aes(colour = PatientID2), size = 3, position = position_dodge(width = 0.75), alpha = 0.5) +
  scale_y_continuous("Minimum infected cell half-life\n(hrs)", limits = c(5, 35)) +
  geom_hline(yintercept = 31.18, alpha = 0.5, linetype = "dotted", size = 1.2) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times"),
        axis.title.y = element_text(size = 11),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank()
  )

peak_box
```

## Time of peak response (when shortest infected cell lifespan)

```{r, results = 'asis'}

print(xtable(
  select(peak_time_table, PatientID2, peak_immune_day),
  digits = 2, caption = "Day of peak immune pressure"),
      comment = F, include.rownames = F)

peak_time_table %>% ungroup() %>%
  summarize(
    median = median(peak_immune_day),
    IQR = paste(round(quantile(peak_immune_day, c(0.25, 0.75)), 3), collapse = ", "),
    range = paste(round(range(peak_immune_day), 3), collapse = ", ")
  ) %>%
  xtable(caption = "Summary stats for day of immune pressure") %>%
  print(comment = F, include.rownames = F)

```

```{r time of peak immune, fig.cap='distribution of peak immune pressure', fig.align='center',fig.height=3,fig.width=3}

peak_immune_time_plot = ggplot(data = immune_summary_data, aes(x = "", y = peak_immune_day)) +
  scale_y_continuous("Days", limits = x_lims) +
  geom_boxplot(width = 1, outlier.colour = alpha("white", 0)) +
  geom_point(aes(colour = PatientID2), size = 3, position = position_dodge(width = 0.75), alpha = 0.5) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times"),
        axis.title.y = element_text(size = 11),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank()
  )

peak_immune_time_plot

```

# Effective reproduction number

## Clearance phase initiation and effective R == 1

When effective R crosses 1 for the first time, that initiates clearance. This matches with the peak viral load day predicted in the model as expected (with some round error)

```{r effR summary statistics, results = 'asis'}
print(xtable(
  select(peak_time_table, PatientID2, peak_dayV, cross_day) %>%
    rename(peak_viral_day = peak_dayV, effectiveR_eq1_day = cross_day), 
  digits = 2, caption = "Day of peak viral load (clearance start) and when R = 1 "),
      comment = F, include.rownames = F)

select(peak_time_table, PatientID2, peak_dayV, cross_day) %>%
    rename(peak_viral_day = peak_dayV, effectiveR_eq1_day = cross_day) %>%
  reshape2::melt(measure.vars = c("peak_viral_day", "effectiveR_eq1_day")) %>%
    group_by(variable) %>%
    summarize(
      median = median(value),
      IQR = paste(round(quantile(value, c(0.25, 0.75)), 3), collapse = ", "),
      range = paste(round(range(value), 3), collapse = ", ")
    ) %>%
    xtable(caption = "Summary stats for day of peak viral load (clearance start) and when R = 1") %>%
    print(comment = F, include.rownames = F)

```

## Comparing clearance phase initiation and peak immune response day

Clearance phase initiation equivalent to first day that R = 1. It happens much earlier than the peak immune pressure day.

```{r blah, fig.cap='distribution of days for critical immune times', fig.align = 'center',fig.height=3.5,fig.width=3.5}

clearance_data = reshape2::melt(immune_summary_data, measure.vars = c("cross_day", "peak_immune_day"))
clr_plot = ggplot(data = clearance_data, aes(x = variable, y = value)) +
  geom_boxplot(width = 0.5, outlier.colour = alpha("white", 0)) +
  geom_point(aes(colour = PatientID2), size = 3, 
             position = position_dodge(width = 0.5), alpha = 0.5) +
  scale_x_discrete(labels = c("Clearance phase\ninitiation", "Maximum immune\npressure")) +
  scale_y_continuous("Days", limits = x_lims) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times"),
        axis.title.y = element_text(size = 11),
        axis.text.x = element_text(size = 8),
        axis.title.x = element_blank()
  )

clr_plot

```


## Minimum effective R vs R0

```{r, results = 'asis'}
R0_data = immune_summary_data %>% group_by(PatientID2, R0, minR) %>%
  reshape2::melt(measure.vars = c("R0", "minR"))

print(xtable(
  select(immune_summary_data, PatientID2, R0, minR),
  digits = 2, caption = "R0 and minimum effective R"),
      comment = F, include.rownames = F)

R0_data %>% group_by(variable) %>%
    summarize(
      median = median(value),
      IQR = paste(round(quantile(value, c(0.25, 0.75)), 3), collapse = ", "),
      range = paste(round(range(value), 3), collapse = ", ")
    ) %>%
    xtable(caption = "Summary stats for R0 and minimum effective R", digits = 3) %>%
    print(comment = F, include.rownames = F)

```


```{r, fig.cap="R0 vs minimum R", fig.align = 'center', fig.height=3.5,fig.width=3.5}
R0_plot = ggplot(data = R0_data, aes(x = variable, y = value)) +
  geom_boxplot(width = 0.5, outlier.colour = alpha("white", 0)) +
  geom_point(aes(colour = PatientID2), size = 3, 
             position = position_dodge(width = 0.5), alpha = 0.5) +
  scale_x_discrete(labels = c(expression(paste(R[0])), "Minimum")) +
  scale_y_continuous("Effective reproduction number", limits = c(0.75, 3.2), breaks = 1:13/2) +
  geom_hline(yintercept = 1, alpha = 0.5, linetype = "dotted", size = 1.2) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times"),
        axis.title.y = element_text(size = 11),
        axis.title.x = element_blank()
  )

R0_plot
```

## Effective R over time

```{r time series R, fig.align = 'center', fig.cap='Immune pressure over time on infected cell lifespan', fig.height = 4, fig.width = 4}
dynR0_plot =  ggplot(data = immune_sim_only, aes(x = days_model_adj, y = log10(effR0), colour = PatientID2)) + 
  geom_line() +
  scale_y_continuous(expression(paste(Log[10], " effective reproduction number")), limits = c(-0.5, 0.5), breaks = -2:2/4) +
  scale_x_continuous("Time (days)", limits = x_lims) +
  geom_hline(yintercept = 1, alpha = 0.5, linetype = "dotted", size = 1.2) +
  scale_colour_discrete(guide = F) +
  theme(legend.position = "none",
        panel.background = element_rect(colour = "black"),
        axis.title.y = element_text(size = 11),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, family = "Times")
  )

dynR0_plot
```

# Immune effect on epithelial cell population

Only one of the model predicts substantial population depletion

```{r, results = 'asis'}

cell_loss = plyr::ldply(unique(immune_model_fits$PatientID2), function(pid){
  p_data_temp = subset(sim_data, PatientID2 == pid)
  data.frame(
    PatientID2 = pid,
    max_loss = max(1 - (p_data_temp$S + p_data_temp$I + p_data_temp$I0)/4e8, na.rm = T)
  )
})
cell_loss %>% xtable(caption = "Maximum cell loss (percent)", digits = 3) %>% print(include.rownames = F, comment = F)

cell_loss %>%
  summarize(
    median_pct_loss = median(max_loss),
    range_pct_loss = paste(round(range(max_loss), 3), collapse = ", ")
  ) %>%
  xtable(caption = "Summary of maximum cell loss (percent)", digits = 3) %>% print(include.rownames = F, comment = F)
```

```{r, fig.cap = 'Population depletion of epithelial cells'}

ggplot(data = subset(sim_data, model != "data"),
       aes(x = days_model, y = ((S + I)/40e7))) + 
  geom_line(aes(colour = PatientID2)) + facet_wrap(~model) +
  scale_y_continuous("Remaining epithelial cells / initial epithelial cells") +
  scale_colour_discrete(guide = F)

```

# Rebound parameters

```{r, results = 'asis'}

rebound_table = readr::read_csv("../tables/supplementary-tables/supptable_cubic_regression.csv")
rebound_table$rebound = with(rebound_table, ifelse(is.na(rebound), "0", rebound))

rebound_parameters = full_join(immune_model_fits, rebound_table, by = "PatientID2")

rebound_parameters$rebound_cat = with(rebound_parameters, ifelse(rebound == "**", "rebound", "no rebound"))
rebound_parameters$rebound_cat2 = with(rebound_parameters, ifelse(rebound == "*", "deceleration", rebound_cat))

rebound_parameters %>% group_by(rebound_cat) %>%
  summarize(
    'median theta' = median(theta),
    'median log10 KI)' = median(log10(KI)),
    'median effector lifespan (1/gamma)' = 1/median(death)
  ) %>%
  xtable(caption = "Immune parameter differences by rebound status (two category)") %>% print(comment = F, include.rownames = F)

rebound_parameters %>% group_by(rebound_cat2) %>%
  summarize(
    'median theta' = median(theta),
    'median log10 KI)' = median(log10(KI)),
    'median effector lifespan (1/gamma)' = 1/median(death)
  ) %>%
  xtable(caption = "Immune parameter differences by rebound status") %>% print(comment = F, include.rownames = F)


```

```{r rebound plots, fig.caption='parameter distributions by rebound'}
rebound_parameters_long = rebound_parameters %>% rename(gamma = death) %>%
  reshape2::melt(measure.vars = c("theta", "KI", "gamma"))
  

rbnd_pl = ggplot(data = rebound_parameters_long, aes(x = variable, y = log10(value), colour = rebound_cat)) +
  geom_boxplot() +
  theme(legend.position = "top", legend.title = element_blank())

rbnd_pl2 = ggplot(data = rebound_parameters_long, aes(x = variable, y = log10(value), colour = rebound_cat2)) +
  geom_boxplot() +
  theme(legend.position = "top", legend.title = element_blank())

plot_grid(rbnd_pl, rbnd_pl2)


```

# Save output data and figures.

See code for this.
```{r}
if(save_output_figs){

  # Figure 3 - no extra modifications
  ggsave("../figures/Figure3.pdf",
     all_plots, width = 210, height = 297, units = "mm")
  

  
  # Figure 4 - combine the component plots
  box_plots_three = plot_grid(peak_box, clr_plot, R0_plot, nrow = 1, labels = LETTERS[2:4], align = "h", hjust = 0)
  plot_dyn =  plot_grid(dyn_plot, labels = "A", hjust = 0)
  plot_dynR02 =  plot_grid(dynR0_plot, labels = "E")
  figure_4 = plot_grid(plot_dyn, box_plots_three, plot_dynR02, nrow = 3, align = "v", hjust = 0)

  ggsave("../figures/Figure4.pdf",
       figure_4, width = 210, height = 200, units = "mm")

}

if(save_output_tables){
  immune_fits_table = select(immune_fits_table, -mse)
  names(immune_fits_table) = c("Infant ID", "theta", "log10 KI", "gamma (x 100)")
  readr::write_csv(immune_fits_table, "../tables/supplementary-tables/immune_model_fits.csv")
}

```

